---
title: "Coding Assignment 3"
subtitle: "CS598: Practical Statistical Learning"
author: 
 - name: Naomi Bhagat - nbhagat3
 - name: Michael Miller - msmille3
 - name: Joe May - jemay3
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: readable
    toc: yes
urlcolor: cyan
---

# Assignment Data
Program: MCS-DS  
Team contributions:

| Person         | Contribution |
|----------      |----------|
| Naomi Bhagat   | Part 1 |
| Michael Miller |  |
| Joe May        |  |

Assignment URL: [campuswire post]()

# Part 1: Optimal Span for LOESS
## Part 1-1: Computing the Diagonal of the Smoother Matrix

Create a function to retrieve the diagonal of the smoother matrix. We’re only
interested in the diagonal entries (which will be used in computing LOO-CV and
GCV), so this function should return an n-by-1 vector.
- Inputs: x (an n-by-1 feature vector) and span (a numerical value).
- Output: n-by-1 vector representing the diagonal of the smoother matrix S.
- Tip: Review the technique we used for the smoother matrix in smoothing spline 
models and adapt it for LOESS.

```{r part1-1, eval = TRUE}
get_diag = function(x, span) {
  # x: n-by-1 feature vector
  # span: numerical value for span
  
  n = length(x)
  
  # return vector d (diagonal)
  d = rep(0, n)
  
  # Smoothing matrix S
  S = diag(n)
  
  # Algorithm
  for(i in 1:n) {
    fit = loess(S[ , i] ~ x, span=span)
    y_hat = fit$fitted
    d[i] = y_hat[i]
  }
  
  return(d)
}
```

### Part 1-2
Span Value Iteration:
- Iterate over the specified span values.
- For each span, calculate the CV and GCV values.
- Post iteration, compile lists of CV and GCV values corresponding to each span.

```{r part1-2, eval = TRUE}
span_value_iter = function(x, y, span) {
  n = length(x)
  spans = length(span)
  
  # return value vectors
  cv = rep(0, spans)
  gcv = rep(0, spans)
  
  # Iterate over the span values
  for(s in 1:spans) {
    # For each span, calculate cv and gcv values
    # First, fit the loess model
    fit = loess(y ~ x, span=s)
    residuals = fit$residuals
    d = get_diag(x, s)
    
    # intermediate calculations for cv and gcv
    # https://liangfgithub.github.io/Notes/lec_W5_NonlinearRegression.pdf slide 33
    tr = sum(d)
    sse = sum(residuals ^ 2)
    
    # calculate cv
    cv[s] = sum((residuals / (1 - d))^2) / n
    
    # calculate gcv
    gcv[s] = sse / (n * (1 - (tr / n)) ^ 2)
  }
  
  return(list(cv = cv, gcv = gcv))
}
```

### Part 1-3
1. Test your code using data set [Coding3_Data.csv]
2. Report your CV and GCV for the following 15 span values: 0.20, 0.25, . . . , 0.90.
3. Report the optimal span value(s) based on CV and GCV.
4. Display the original data points and overlay them with the true curve and the fitted curve(s) generated using the optimal span value(s).

The true curve is \[ f(x) = \frac{\sin(12(x + 0.2))}{x + 0.2} \]

```{r part1-3, eval = TRUE}
part1_data = read.csv("Coding3_Data.csv")
test_span = c(0.2, 0.25, 0.3, 0.35, 0.4, 0.45, 0.5, 0.55, 0.6, 0.65, 0.7, 0.75, 0.8, 0.85, 0.9)
result = span_value_iter(part1_data$x, part1_data$y, test_span)

# Report cv, gcv values
output = data.frame(CV = result$cv, GCV = result$gcv, span = test_span)
output

# Optimal cv values
cv_opt = output$span[output$CV == min(output$CV)]
cv_opt

# Optimal gcv values
gcv_opt = output$span[output$GCV == min(output$GCV)]
gcv_opt

# Plot curve
# Data + Control curve
plot(part1_data$x, part1_data$y, xlab="x", ylab="y")

fx = 1:50/50
fy = sin(12*(fx + 0.2)) / (fx + 0.2)
lines(fx, fy, col="red")

# My curve - cv
cv_f = loess(y ~ x, part1_data, span=cv_opt)
lines(fx, predict(cv_f, data.frame(x = fx)), lty=2, col="blue")

# My curve - gcv
gcv_f = loess(y ~ x, part1_data, span=gcv_opt)
lines(fx, predict(gcv_f, data.frame(x = fx)), lty=2, col="green")
```

