---
title: "Coding Assignment 3"
subtitle: "CS598: Practical Statistical Learning"
author: 
 - name: Naomi Bhagat - nbhagat3
 - name: Michael Miller - msmille3
 - name: Joe May - jemay3
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: readable
    toc: yes
urlcolor: cyan
---

# Assignment Data
Program: MCS-DS  
Team contributions:

| Person         | Contribution |
|----------      |----------|
| Naomi Bhagat   | Part 1 |
| Michael Miller | Part 2 |
| Joe May        |  |

Assignment URL: [campuswire post](https://campuswire.com/c/G06C55090/feed/242)

```{r setup, eval = TRUE}
# add required packages
packages = c('splines')

# if packages don't exist, install. Then call library on them
for (package in packages) {
  if (!requireNamespace(package, quietly=TRUE)) {
    install.packages(package)
  }
  library(package, character.only=TRUE)
}

# set seed
set.seed(235)

```


# Part 1: Optimal Span for LOESS
## Part 1-1: Computing the Diagonal of the Smoother Matrix

Create a function to retrieve the diagonal of the smoother matrix. We’re only
interested in the diagonal entries (which will be used in computing LOO-CV and
GCV), so this function should return an n-by-1 vector.
- Inputs: x (an n-by-1 feature vector) and span (a numerical value).
- Output: n-by-1 vector representing the diagonal of the smoother matrix S.
- Tip: Review the technique we used for the smoother matrix in smoothing spline 
models and adapt it for LOESS.

```{r part1-1, eval = TRUE}
get_diag = function(x, span) {
  # x: n-by-1 feature vector
  # span: numerical value for span
  
  n = length(x)
  
  # return vector d (diagonal)
  d = rep(0, n)
  
  # Smoothing matrix S
  S = diag(n)
  
  # Algorithm
  for(i in 1:n) {
    fit = loess(S[ , i] ~ x, span=span)
    y_hat = fit$fitted
    d[i] = y_hat[i]
  }
  
  return(d)
}
```

### Part 1-2
Span Value Iteration:
- Iterate over the specified span values.
- For each span, calculate the CV and GCV values.
- Post iteration, compile lists of CV and GCV values corresponding to each span.

```{r part1-2, eval = TRUE}
span_value_iter = function(x, y, span) {
  n = length(x)
  num_span = length(span)
  
  # return value vectors
  cv = rep(0, num_span)
  gcv = rep(0, num_span)
  
  # Iterate over the span values
  for(s in 1:num_span) {
    # For each span, calculate cv and gcv values
    # First, fit the loess model
    fit = loess(y ~ x, span=span[s], control=loess.control(surface="direct"))
    residuals = fit$residuals
    d = get_diag(x, span[s])
    
    # intermediate calculations for cv and gcv
    # https://liangfgithub.github.io/Notes/lec_W5_NonlinearRegression.pdf slide 33
    tr = sum(d)
    sse = sum(residuals ^ 2)
    
    # calculate cv
    cv[s] = sum((residuals / (1 - d))^2) / n
    
    # calculate gcv
    gcv[s] = sse / (n * (1 - (tr / n)) ^ 2)
  }
  
  return(list(cv = cv, gcv = gcv))
}
```

### Part 1-3
1. Test your code using data set [Coding3_Data.csv]
2. Report your CV and GCV for the following 15 span values: 0.20, 0.25, . . . , 0.90.
3. Report the optimal span value(s) based on CV and GCV.
4. Display the original data points and overlay them with the true curve and the fitted curve(s) generated using the optimal span value(s).

The true curve is \[ f(x) = \frac{\sin(12(x + 0.2))}{x + 0.2} \]

```{r part1-3, eval = TRUE}
part1_data = read.csv("Coding3_Data.csv")
test_span = c(0.2, 0.25, 0.3, 0.35, 0.4, 0.45, 0.5, 0.55, 0.6, 0.65, 0.7, 0.75, 0.8, 0.85, 0.9)
result = span_value_iter(part1_data$x, part1_data$y, test_span)

# Report cv, gcv values
output = data.frame(CV = result$cv, GCV = result$gcv, span = test_span)
output

# Optimal cv values
cv_opt = output$span[output$CV == min(output$CV)]
cv_opt

# Optimal gcv values
gcv_opt = output$span[output$GCV == min(output$GCV)]
gcv_opt

# Plot curve
# Data + Control curve
plot(part1_data$x, part1_data$y, xlab="x", ylab="y")

fx = 1:50/50
fy = sin(12*(fx + 0.2)) / (fx + 0.2)
lines(fx, fy, col="red")

# My curve - cv
cv_f = loess(y ~ x, part1_data, span=cv_opt)
lines(fx, predict(cv_f, data.frame(x = fx)), lty=2, col="green")

# My curve - gcv
gcv_f = loess(y ~ x, part1_data, span=gcv_opt)
lines(fx, predict(gcv_f, data.frame(x = fx)), lty=2, col="blue")
```

# Part 2: Clustering Time Series
## Preparation

```{r part2-0, eval = TRUE}
sales_df = read.csv('Sales_Transactions_Dataset_Weekly.csv')

# Per https://campuswire.com/c/G06C55090/feed/346, only keep W columns
sales_df = subset(sales_df, select = grepl('W', names(sales_df)))

#de-mean
sales_df = sales_df - rowMeans(sales_df)

X = as.matrix(sales_df)
```

## Part 2-1: Fitting NCS
- Fit each time series with an NCS with df = 10. This corresponds to an NCS with 8 interior knots. Each row of X represents the response, with the 1-dimensional feature being the index from 1 to 52.
- Store the NCS coefficients (excluding the intercept) in an 811-by-9 matrix B

```{r part2-1, eval = TRUE}
# initialize constants for later use
weeks = 1:52
df = 10

# get design matrix and remove intercept 
f = as.matrix(ns(weeks, df=df-1, intercept=FALSE))

# get matrix of splines
B = t(solve(t(f) %*% f) %*% t(f) %*% t(X))
```


Part 2-2: Clustering using Matrix B
```{r part2-2, eval = TRUE}
#6 clusters per instructions 
cluster_count = 6

# get clusters
clusters = kmeans(B, centers=cluster_count)

# initialize new plot with 6 windows, add some margin for title
par(mfrow = c(2, 3), oma = c(0, 0, 4, 0))

# iterate through clusters, plot each one
for (cluster in 1:cluster_count) {
  # get data for this cluster
  x_cluster = X[clusters$cluster == cluster, ]
  b_cluster = B[clusters$cluster == cluster, ]
  
  # convert b spline back to timeseries
  b = colMeans(b_cluster)
  spline_plot = f %*% b
  
  plot(x=weeks, 
       y=spline_plot,
       main=paste('Cluster ', cluster),
       ylab='Weekly Sales (delta)',
       xlab='Week',
       ylim=c(-20, 30),
       col='blue',
       type = 'l')
  
  # plot each x in this cluster
  for (series in 1:nrow(x_cluster)) {
    lines(x=weeks, 
          y=x_cluster[series, ],
          col="gray")
  }
  
  # now plot the cluster center again for visibility (so its 'on top')
  lines(x=weeks, 
        y=spline_plot,
        col="blue",
        lwd=2)
}

# add main title to top
mtext('Custered Timeserires Data with Spline Centers (B)', side=3, line=1, outer=TRUE)

```

Part 2-3: Clustering using Matrix X 
```{r part2-3, eval = TRUE}
# get clusters
clusters = kmeans(X, centers=cluster_count)

# initialize new plot with 6 windows, add some margin for title
par(mfrow = c(2, 3), oma = c(0, 0, 4, 0))

# iterate through clusters, plot each one
for (cluster in 1:cluster_count) {
  # get data for this cluster
  x_cluster = X[clusters$cluster == cluster, ]
  
  # this time spline plot is mean of clusters
  center_plot = colMeans(x_cluster)
  
  plot(x=weeks, 
       y=center_plot,
       main=paste('Cluster ', cluster),
       ylab='Weekly Sales (delta)',
       xlab='Week',
       ylim=c(-20, 30),
       col='blue',
       type = 'l')
  
  # plot each x in this cluster
  for (series in 1:nrow(x_cluster)) {
    lines(x=weeks, 
          y=x_cluster[series, ],
          col="grey")
  }
  
  # now plot the cluster center again for visibility (so its 'on top')
  lines(x=weeks, 
        y=center_plot,
        col="blue",
        lwd=2)
  
# add main title to top
mtext('Custered Timeserires Data with Mean Centers (X)', side=3, line=1, outer=TRUE)

}
```


# Part 3: Ridgeless and double descent
## Part 3-1: Ridgeless Function

```{r part3-1, eval = TRUE}

part3_data = read.csv('Coding3_dataH.csv',header = FALSE)

data = as.matrix(part3_data)


ridgeless = function(train, test){
trainX = train[,-1]
trainY = train[,1]
testX = test[,-1]
testY = test[,1]

pca = prcomp(trainX[,-1], scale = FALSE, center = TRUE)
pc = pca$rotation

}



```

## Part 3-2: Simulation study
Graphical display: Plot the median of the test errors (collated over the 30 iterations) in log scale against the count of regression parameters, which spans from 5 to 240.

```{r part3-2, eval = TRUE}

# placeholder for testing
ridgeless = function(train, test){
  return(runif(nrow(test), min = 0, max = ncol(test)))
}


# params for simulation 
n_sims = 30
p_train = .75
n_train = round(nrow(data) * p_train)
d_vals = 6:241

# preallocate data collection
errors = matrix(0, nrow=n_sims, ncol=length(d_vals))

#iterate, run each simulation
for (sim in 1:n_sims) {
  # partition data
  train_idx = sample(1:nrow(data), n_train)
  train = data[train_idx, ]
  test = data[-train_idx, ]
  
  # iterate through values of d
  for (d in 1:length(d_vals)) {
    # subset data to only account for first d params
    train_subset = train[ ,1:d_vals[d]]
    test_subset = test[ ,1:d_vals[d]]
    
    # this part depends a bit on specific implementation of ridgeless
    # error needs to be vector of test error
    out = ridgeless(train_subset, test_subset)
    error = log(sum(out))
    
    errors[sim, d] = error
  }
}

# average across simulations
sim_output = apply(errors, 2, median)

plot(x=d_vals-1,
     y=sim_output, 
     main=paste('Simulation Study: Error vs. Feature Count'),
     ylab='Log(TestError)',
     xlab='Feature Count',
     log='y',
     col='blue'
     )

```
