---
title: "Project 4: Movie Recommender System"
subtitle: "CS598: Practical Statistical Learning"
author: 
 - name: Naomi Bhagat - nbhagat3
 - name: Michael Miller - msmille3
 - name: Joe May - jemay3
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: readable
    toc: yes
urlcolor: cyan
---

# Assignment Data
Program: MCS-DS  
Team contributions:

| Person         | Contribution |
|----------      |----------|
| Naomi Bhagat   | Application |
| Michael Miller | System 1 |
| Joe May        | System 2  |

Assignment URL: [campuswire post](https://liangfgithub.github.io/Proj/F23_Proj4.html)

```{r setup, eval = TRUE}
# add required packages
packages = c('')

# if packages don't exist, install. Then call library on them
for (package in packages) {
  if (!requireNamespace(package, quietly=TRUE)) {
    install.packages(package)
  }
  library(package, character.only=TRUE)
}

# set seed
set.seed(235)

```


# System I: Recommendation Based on Genres

We propose the following genre-based recommendation system. A human-readable description of the system is provided first, with implementation in r below. This section of the report generates a csv file that will be used directly by the application to make recommendation.


## Considerations
Goals 
Show users movies that are good
Get more reviews 


1. Recommend 'good' movies. Define good 

2. Add in some highly rated by not very well reviewed movies

3. Add in a small number of not very well reviewed movies 


## Limitations
Algorithm implemented without feedback
Algorithm is static

In real world setting, would try to get additional reviews from users. Would recalculate recommendations each time so that users see new recommendations. A good strategy might be: once a user 'selects' a movie, to replace it with a new recommendation, or provide some kind of shuffle button for the user to reset recommendations. 

# Implementation 
```{r ingest_data, eval = TRUE}
# download and preprocses raw movie data
# note that this code is from example code from this class 
# https://liangfgithub.github.io/Rcode_W13_Movie_EDA.nb.html

source_url = "https://liangfgithub.github.io/MovieData/"

ratings = read.csv(paste0(source_url, 'ratings.dat?raw=true'), 
                   sep = ':',
                   colClasses = c('integer', 'NULL'), 
                   header = FALSE)

colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')

# load movies. Same source
movies = readLines(paste0(source_url, 'movies.dat?raw=true'))
movies = strsplit(movies, split = "::", fixed = TRUE, useBytes = TRUE)
movies = matrix(unlist(movies), ncol = 3, byrow = TRUE)
movies = data.frame(movies, stringsAsFactors = FALSE)
colnames(movies) = c('MovieID', 'Title', 'Genres')
movies$MovieID = as.integer(movies$MovieID)

# convert accented characters
movies$Title = iconv(movies$Title, "latin1", "UTF-8")

# extract year
movies$Year = as.numeric(unlist(
  lapply(movies$Title, function(x) substr(x, nchar(x)-4, nchar(x)-1))))

# 
users = read.csv(paste0(source_url, 'users.dat?raw=true'),
                 sep = ':', header = FALSE)
users = users[, -c(2,4,6,8)] # skip columns
colnames(users) = c('UserID', 'Gender', 'Age', 'Occupation', 'Zip-code')

```

# System II: Recommendation Based on IBCF

```{r system2, eval = TRUE}
```

