---
title: "Coding Assignment 5"
subtitle: "CS598: Practical Statistical Learning"
author: 
 - name: Naomi Bhagat - nbhagat3
 - name: Michael Miller - msmille3
 - name: Joe May - jemay3
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: readable
    toc: yes
urlcolor: cyan
---

# Assignment Data
Program: MCS-DS  
Team contributions:

| Person         | Contribution |
|----------      |----------|
| Naomi Bhagat   | Evaluation |
| Michael Miller | Pegasos implementation |
| Joe May        | Fine tuning |

Assignment URL: [campuswire post](https://campuswire.com/c/G06C55090/feed/602)

```{r setup, eval = TRUE}
# set seed
set.seed(235)

```

# Part 1: Linear SVM 
## Part 1-1: Implement the Pegasos Algorithm  
Use a fixed number of epochs, e.g., T = 20.  
In each epoch, before going through the dataset, consider randomizing the order of the data points. To achieve this, you should set random seeds for shuffling. For this assignment, the seeds used for shuffling do not need to be associated with your UIN.

```{r part1-1, eval = TRUE}

# helper function for getting predictions from params
get_predictions = function(X, B, a){
  out = as.matrix(X) %*% as.matrix(B) + a
  
  out_full = ifelse(out > 0, 1, -1)
  
  return(out_full)
}

pegasos = function(train, epochs=20, lambda=1) {
  # get X and Y as a matrix
  X = as.matrix(train[ ,-ncol(train)])
  Y = train[ ,ncol(train)]
  
  n = nrow(X)
  m = ncol(X)
  
  # initialize params
  B = rep(0, m)
  a = 0
  t = 0
  
  for(epoch in 1:epochs) {
    # shuffle input
    sample_idx = sample(n)
    
    shuffle_x = X[sample_idx, ]
    shuffle_y = Y[sample_idx]
    
    # iterate through points
    for (i in 1:n) {
      # get x and y
      x = shuffle_x[i ,]
      y = shuffle_y[i]
      
      # this term is used a bunch below
      eval_term = y * (sum(x * B) + a)
      
      # calc subgradient terms delta t and del t
      if (eval_term < 1) {
        delta_t = lambda * B - y*x
        del_t = -y
      } else {
        delta_t = lambda*B
        del_t = 0
      }
      
      # update params 
      t = t + 1
      eta = 1 / (t * lambda)
      B = B - eta * delta_t
      a = a - eta * del_t
    }
    
    # print current status
    current_pred = get_predictions(X, B, a)
    error_fraction = sum(current_pred != Y) / length(Y)
    print(paste0("Epoch: ", epoch, ". Error fraction: ", error_fraction))
  }
  
  # return out latest a and B
  return(list(a=a, B=B))
}
```


## Part 1-2: Training 
Test your code with the provided training (200 samples) and test (600 samples) datasets, which are subsets of the MNIST data. Each dataset consists of 257 columns, with the first 256 columns representing the features, and the last column indicating the label (either 5 or 6).  
Report confusion tables on the training and test datasets.  
Your code should obtain less than 15% test error.  

```{r part1-2, eval = TRUE}
  # read in data
  train = read.csv('coding5_train.csv')
  test = read.csv('coding5_test.csv')
  
  # set aside y with values still set to 5 and 6
  train_y_actual = train$Y
  test_y_actual = test$Y
  
  # for training, preprocess  so that y is -1 or 1
  train$Y = ifelse(train$Y == 6, 1, -1)
  
  # train SVM
  model = pegasos(train, lambda=.01, epochs=10)
```

## Part 1-3: Evaluation 
Report confusion tables on the training and test datasets.  
Your code should obtain less than 15% test error.  

```{r part1-3, eval = TRUE}
  # get predictions 
  train_y_pred = get_predictions(train[ ,-ncol(train)], model$B, model$a)
  test_y_pred = get_predictions(test[ ,-ncol(test)], model$B, model$a)
  
  # post process predictions back to 5 and 6
  train_y_pred = ifelse(train_y_pred == 1, 6, 5)
  test_y_pred = ifelse(test_y_pred == 1, 6, 5)


  train_error_percent = round(sum(train_y_pred != train_y_actual)/length(train_y_actual) * 100, 2)
  test_error_percent = round(sum(test_y_pred != test_y_actual)/length(test_y_actual) * 100, 2)
```
Train error: `r train_error_percent`%  
Test error: `r test_error_percent`%

Confusion table for train dataset: 
```{r part1-4, echo=FALSE}
  table(Actual=train_y_actual, Predicted=train_y_pred)
```

Confusion table for test dataset: 
```{r part1-5, echo=FALSE}
  table(Actual=test_y_actual, Predicted=test_y_pred)
```
